---
Authors: ["**Achal Neupane**"]
title: "Graphical Summary"
date: 2019-10-01T17:26:23-05:00
draft: false
output: html_document
tags:
- R
- Statistics
- Modern Applied Statistics I
summary: Statistics series
---

<script type="text/javascript" src="https://cdnjs.cloudflare.com/ajax/libs/mathjax/2.7.1/MathJax.js?config=TeX-AMS-MML_HTMLorMML">
</script>



Answer all questions specified on the problem and include a discussion on how your results answered/addressed the question.

Submit your \textbf{.rmd} file with the knitted \textbf{PDF} (or knitted Word Document saved as a PDF). If you are having trouble with .rmd, let us know and we will help you, but both the .rmd and the PDF are required.

This file can be used as a skeleton document for your code/write up. Please follow the instructions found under Content titled Format+STAT-701+HW. No code should be in your PDF write-up unless stated otherwise.

For any question asking for plots/graphs, please do as the question asks as well as do the same but using the respective commands in the GGPLOT2 library. (So if the question asks for one plot, your results should have two plots. One produced using the given R-function and one produced from the GGPLOT2 equivalent). This doesn't apply to questions that don't specifically ask for a plot, however I still would encourage you to produce both.

You do not need to include the above statements.

Please do the following problems from the text book R Handbook and stated.

1. Question 1.1, pg. 23 in Handbook \textit{this question will require you to make some assumptions. List your assumptions and how you interpreted the question.}

Here, for this question, we assume that we have data available for the given countries and we will remove any NAs from the data.

```{r}
library('HSAUR3')
#subset the data for wanted contries only first
median.subset <- subset(Forbes2000,country %in% c("United States","United Kingdom","France","Germany"))
# library("plyr")
# ddply(median.subset,c("country"),function(x){median(x$profits,na.rm=T)})
by.country <- setNames(aggregate(median.subset$profits, by = list(median.subset$country), function(x){median(x,na.rm=T)}), c("country", "median"))
by.country
```
2. Question 1.2, pg. 23 in Handbook

```{r}
Forbes2000[Forbes2000$country == "Germany" & Forbes2000$profits < 0,"name"]
```
3. Question 1.3, pg. 23 in Handbook

```{r}
sort(table(Forbes2000[Forbes2000$country=="Bermuda","category"]),decreasing=T)
```

4. Question 1.4, pg. 23 in Handbook


```{r}

Forbes2000.data <- Forbes2000[order(Forbes2000$profits,decreasing=T),]
par(mfrow = c(1,1))
# Using base plot
plot(
    log(Forbes2000.data[1:50, "assets"]),
    log(Forbes2000.data[1:50, "sales"])
    ,
    ylab = "sales (log scale)",
    xlab = "assets (log scale)",
    main = "Sales vs Assets: Log transformed"
)
# we can add texts to data points
text(
    x = log(Forbes2000.data[1:50, "assets"]),
    y = log(Forbes2000.data[1:50, "sales"]),
    labels = abbreviate(Forbes2000.data$country[1:50]),
    col = "black",
    font = 1,
    pos = 1
)

#Using ggplot
library(ggplot2)
ggplot(Forbes2000.data[1:50, ], aes(x = log(assets), y = log(sales))) +
    geom_point(shape = 1, size = 2) +
    geom_text(aes(label = abbreviate(country)), hjust = 0, vjust = 0) +
    labs(title = "Sales vs Assets: Log transformed", color = "country")
```


5. Question 1.5, pg. 23 in Handbook

```{r}
library(plyr)
# selecting companies with profit more than 5 billion
Forbes2000.5billions.dat <- Forbes2000[Forbes2000$profits > 5 ,]
Forbes2000.5billions.dat <- Forbes2000.5billions.dat[complete.cases(Forbes2000.5billions.dat), ]

setNames(ddply(Forbes2000.5billions.dat,c("country"),function(x){c(nrow(x),mean(x$sales))}), c("Country", "Company_counts", "Avg_Sales"))

# #or we could also do:
# aggregate(Forbes2000.5billions.dat$sales, by = list(Forbes2000.5billions.dat$country), function(x){c(Company_counts = length(x), Avg_Sales = mean(x))})
```
6. Question 2.1, pg. 39 in Handbook (see Chapter 6 of R Graphcis Cookbook for GGPlot)

For this, we will first calculate total household expenses as below. Then calculate the itemized expenses (proportion). 

```{r}
data("household", package = "HSAUR3")

# total expenses
household$total<-household$housing+household$food+household$goods+household$service
library(ggplot2)

ggplot(household, aes(x=gender, y=total))+geom_boxplot() +
  labs(title = "Total expenses per gender") + 
  theme_classic()

# base R version
boxplot(household$total ~ household$gender,
        xlab = "Gender", ylab = "total", main = "Total household expenses by gender")

# This indicates that males tend to have more expenses than females in total. 
#calculate proportions
household$housing<-household$housing/household$total
household$service<-household$service/household$total
household$goods<-household$goods/household$total
household$food<-household$food/household$total

#plot boxplots of proportions
a<-ggplot(household, aes(x=gender, y=housing))+geom_boxplot() + labs(title = "Housing expenses (in %) per gender") 
b<-ggplot(household, aes(x=gender, y=service))+geom_boxplot() + labs(title = "Service expenses (in %) per gender")
c<-ggplot(household, aes(x=gender, y=goods))+geom_boxplot() + labs(title = "Goods expenses (in %) per gender")
d<-ggplot(household, aes(x=gender, y=food))+geom_boxplot() + labs(title = "Food expenses (in %) per gender")

# combine all 4 gg objects
library(gridExtra)
grid.arrange(a,b,c,d, nrow=2, ncol=2)

# base R version
par(mfrow=c(2,2))
boxplot(household$housing ~ household$gender,
        xlab = "Gender", ylab = "Housing expenses", main = "Housing expenses (in %) by gender")
boxplot(household$service ~ household$gender,
        xlab = "Gender", ylab = "Service expenses", main = "Service expenses (in %) by gender")
boxplot(household$goods ~ household$gender,
        xlab = "Gender", ylab = "Goods expenses", main = "Goods expenses (in %) by gender")
boxplot(household$food ~ household$gender,
        xlab = "Gender", ylab = "food expenses", main = "Food expenses (in %) by gender")

# histogram of total expenditures by gender
ggplot(household, aes(x=total, fill=gender)) +
  geom_histogram(position="identity", alpha=0.4, binwidth = 1000) + 
  labs(title = "Histogram of total expenditures by gender" ) + 
  theme_classic()
```

  Based on these results, males tend to spend more money than females on Foods
  and Services, where as females take the lead on Housing and Goods.

7. Question 2.3, pg. 41 in Handbook (see Chapter 6 of R Graphcis Cookbook for GGPlot)

```{r}
data("suicides2")
#Boxplot of mortality
boxplot(suicides2,
        xlab = "Age group", ylab = "Mortality rates per 100,000", main = "Mortality by suicide")

library(reshape2)
suicides2.melted <- melt(suicides2)
# ggplot version
ggplot(suicides2.melted, aes(x=factor(variable), y=value)) + 
  geom_boxplot() +
  labs(x = "Age group", y = "Mortality rates per 100,000", title = "Mortality by suicide")
```

8. Using a single R expression, calculate the median absolute deviation, $1.4826\cdot median|x-\mu|$, where $\mu$ is the sample median. Use the dataset \textbf{chickwts}. Use the R function mad() to verify your answer.

```{r}
median(abs(chickwts$weight - median(chickwts$weight))) * 1.4826

#check using mad
mad(chickwts$weight, median (chickwts$weight), constant = 1.4826)
```

10. Using the data matrix \textbf{state.x77}, find the state with the minimum per capita income in the New England region as defined by the factor \textit{state.division}. Use the vector \textit{state.name} to get the state name.

```{r}
data(state.x77)
#create df with state name, income, and division
state_df <- data.frame(income=state.x77[,"Income"],
                    name=state.name,
                    div=state.division)
#Subset for New England
New_Eng <- subset(state_df, div =="New England")

#State with min income per capita
New_Eng[which(New_Eng$income == min(New_Eng$income)), ]
```

11. Use subscripting operations on the dataset \textbf{Cars93} to find the vehicles with highway mileage of less than 25 miles per gallon (variable \textit{MPG.highway}) and weight (variable \textit{Weight}) over 3500lbs. Print the model name, the price range (low, high), highway mileage, and the weight of the cars that satisfy these conditions.

```{r}
data(Cars93, package= "MASS")
df.cars93 <- Cars93[Cars93$MPG.highway < 25 & Cars93$Weight > 3500, c("Model", "Price", "MPG.highway", "Weight")]
df.cars93[with(df.cars93, order(Price)), ]
```

12. Form a matrix object named \textbf{mycars} from the variables \textit{Min.Price, Max.Price, MPG.city, MPG.highway, EngineSize, Length, Weight} from the \textbf{Cars93} dataframe from the \textbf{MASS} package. Use it to create a list object named \textit{cars.stats} containing named components as follows:

    a) A vector of means, named \textit{Cars.Means}
    
    b) A vector of standard errors of the means, named \textit{Cars.Std.Errors}
    
```{r}
data(Cars93, package= "MASS")
mycars <- Cars93[,c("Min.Price", "Max.Price", "MPG.city", "MPG.highway", "EngineSize", "Length", "Weight")]
Cars.Means <- sapply(mycars, mean)       
Cars.Means

# function to calculate standard error 'standard.err'
standard.err <- function(x) sqrt(var(x)/length(x))
Cars.Std.Errors <- lapply(mycars, standard.err)  
Cars.Std.Errors
```
    
13. Use the \texttt{apply()} function on the three-dimensional array \textbf{iris3} to compute:

    a) Sample means of the variables \textit{Sepal Length, Sepal Width, Petal Length, Petal Width}, for each of the three species \textit{Setosa, Versicolor, Virginica}

```{r}
apply(iris3, c(2,3), mean)
```
    
    b) Sample means of the variables \textit{Sepal Length, Sepal Width, Petal Width} for the entire data set.
    
```{r}
apply(iris3, 2, mean)
```
    
14. Use the data matrix \textbf{state.x77} and the \texttt{tapply()} function to obtain:

    a) The mean per capita income of the states in each of the four regions defined by the factor \textit{state.region}
    
```{r}
mydf <- data.frame(state.x77, state.region = state.region, stringsAsFactors = FALSE)
tapply(mydf$Income, mydf$state.region, mean)
```
    b) The maximum illiteracy rates for states in each of the nine divisions defined by the factor \textit{state.division}
    
```{r}
mydf <- data.frame(state.x77, state.division = state.division, stringsAsFactors = FALSE)
tapply(mydf$Illiteracy, mydf$state.division, max)
```
    
    c) The number of states in each region
    
```{r}
mydf <- data.frame(state.x77, state.region = state.region, state.name = state.name, stringsAsFactors = FALSE)
tapply(mydf$state.name, mydf$state.region, length)
```
    
15. Using the dataframe \textbf{mtcars}, produce a scatter plot matrix of the variables \textit{mpg, disp, hp, drat, qsec}. Use different colors to identify cars belonging to each of the categories defined by the \textit{carsize} variable in different colors.

    ```{r, eval=F, echo = T}
    carsize = cut(mtcars[,"wt"], breaks=c(0, 2.5, 3.5, 5.5), 
    labels = c("Compact","Midsize","Large"))
    carsize = cut(mtcars[,"wt"], breaks=c(0, 2.5, 3.5,
                                      5.5), labels = c("Compact","Midsize","Large"))

    mydf <- data.frame(mtcars, carsize = carsize)

    # Using base R
    par(mfrow = c(1,1))
    pairs(~mpg + disp + hp + drat + qsec, data=mydf, 
      main="mtcars Scatterplot Matrix")
    
    # Using ggplot
    library(ggplot2)
    library('GGally')
    ggpairs(mydf[,c("mpg", "disp", "hp", "drat", "qsec")])
    
    ```
    
16. Use the function \texttt{aov()} to perform a one-way analysis of variance on the \textbf{chickwts} data with \textit{feed} as the treatment factor. Assign the result to an object named \textit{chick.aov} and use it to print an ANOVA table.

```{r}
chick.aov <- aov( weight ~ feed, chickwts)

# summary aov
summary.aov(chick.aov)

# post-hoc test (Tukey HSD)
TukeyHSD(chick.aov)
```

17. Write an R function named \texttt{ttest()} for conducting a one-sample t-test. Return a list object containing the two components: 

    - the t-statistic named T;
    
    - the two-sided p-value named P.
    
```{r}    
#Annotated R Code
  library(HSAUR3)
  library(stats)
  attach(chickwts)
  head(chickwts)
  
  t.test(chickwts$weight,mu=240)
  
  ttest=function(x,mu,alpha){
    # x = data
    # mu = sample mean, 
    # alpha = alpha 
    # level = (1-confidence level)
    me=mean(x$weight)
    p1=qt(alpha/2,(nrow(x)-1))
    p2=qt(1-alpha/2,(nrow(x)-1))
    s=sqrt(var(x$weight))
    n=nrow(x)
    
    T=(me-mu)/(s/sqrt(nrow(x))) 
    P=seq(1,1,1)
    P[1]=2*(1-pt(T,n))
    P=data.frame(P)
    return (cbind(P,T))
  }
  
  t_test_values <- ttest(chickwts,240,0.05)
  print ("T value and two sided P values returned by the funtion: ")
  print(t_test_values)
```
    
Use this function to test the hypothesis that the mean of the \textit{weight} variable (in the \textbf{chickwts} dataset) is equal to 240 against the two-sided alternative. \textit{For this problem, please show the code of function you created as well as show the output. }

```{r, message = TRUE}
  message("Hypothesis Result:")
  if (2*abs(t_test_values[1,2])>2*abs(t_test_values[1,1])){
    print("Rejected! The true mean is NOT 240 !!")
  } else if (2*abs(t_test_values[1,2])<2*abs(t_test_values[1,1])){
    print("Null. The mean is 240 !!")
  }
```
    