---
Authors: ["**Achal Neupane**"]
title: "Longitudinal Data Analysis and Mixed Models I"
date: 2019-10-28T17:26:23-05:00
draft: false
output: html_document
tags:
- R
- Statistics
- Modern Applied Statistics I
summary: Statistics series
---

<script type="text/javascript" src="https://cdnjs.cloudflare.com/ajax/libs/mathjax/2.7.1/MathJax.js?config=TeX-AMS-MML_HTMLorMML">
</script>

Answer all questions specified on the problem and include a discussion on how your results answered/addressed the question.

Submit your \textbf{.rmd} file with the knitted \textbf{PDF} (or knitted Word Document saved as a PDF). If you are having trouble with .rmd, let us know and we will help you, but both the .rmd and the PDF are required.

This file can be used as a skeleton document for your code/write up. Please follow the instructions found under Content for Formatting and Guidelines. No code should be in your PDF write-up unless stated otherwise.

For any question asking for plots/graphs, please do as the question asks as well as do the same but using the respective commands in the GGPLOT2 library. (So if the question asks for one plot, your results should have two plots. One produced using the given R-function and one produced from the GGPLOT2 equivalent). This doesn't apply to questions that don't specifically ask for a plot, however I still would encourage you to produce both.

You do not need to include the above statements.

Please do the following problems from the text book R Handbook and stated.


1. Following up with the Beat the Blues data from the video (package HSAUR3) do the following

      a. Construct boxplots to compare the factor variable \textbf{drug} in an analogous way to how we constructed boxplots in the video for the treatment variable. Discuss the results. 

      b. Repeat (a) for the \textbf{length} variable. Discuss the results.

      c. Use the \textit{lm} function to fit a model to the Beat the Blues data that assumes that the repeated measurements are independent. Compare the results to those from fitting the random intercept model \textit{BtheB\_lmer1} from the video.

      d. Investigate and discuss whether there is any evidence of an interaction between treatment and time for the Beat the Blues data.

      e. Construct a plot of the mean profiles of both treatment groups in the Beat the Blues data, showing also standard deviation bars at each time point.
      
      
```{r}
library("gee")
library("lme4")
library("Matrix")
library("multcomp")
library (HSAUR3)
library(ggplot2)
library(plyr)
library(tidyr)

cat("# 1.a") 

# a. Construct boxplots to compare the factor variable \textbf{drug} in an
# analogous way to how we constructed boxplots in the video for the treatment
# variable. Discuss the results.
data("BtheB")

layout(matrix(1:2, nrow=1))

ylim=range(BtheB[ , grep("bdi", names(BtheB))],
           na.rm=TRUE)

yes=subset(BtheB, drug=="Yes")[ , grep("bdi", names(BtheB))]
boxplot(yes, main="Drugs Used", ylab="BDI",
        xlab="Time (in Months)", names=c(0,2,3,5,8), 
        ylim=ylim)

no=subset(BtheB, drug=="No")[ , grep("bdi", names(BtheB))]
boxplot(no, main="Drugs Not Used", ylab="BDI",
        xlab="Time (in Months)", names=c(0,2,3,5,8), 
        ylim=ylim)


cat("#1.b")
# b. Repeat (a) for the \textbf{length} variable. Discuss the results.

layout(matrix(1:2, nrow = 1))
lengthlow<-subset(BtheB,length=="<6m")[,grep("bdi",names(BtheB))]
boxplot(lengthlow, main ="Length<6m", ylab= "BDI", xlab = "Time (in months)",names=c(0,2,3,5,8))

lengthhigh<-subset(BtheB,length==">6m")[,grep("bdi",names(BtheB))]
boxplot(lengthhigh, main ="Length>6m", ylab= "BDI", xlab = "Time (in months)",names=c(0,2,3,5,8))


cat("#1.c")
# c. Use the \textit{lm} function to fit a model to the Beat the Blues
# data that assumes that the repeated measurements are independent.
# Compare the results to those from fitting the random intercept model
# \textit{BtheB\_lmer1} from the video.

BtheB$subject=factor(rownames(BtheB))
nobs=nrow(BtheB)
cat("Let change the data in wide format")
BtheB_long=reshape(BtheB, idvar = "subject",
                   varying = c("bdi.2m", "bdi.3m", "bdi.5m", "bdi.8m"),
                   direction = "long")
BtheB_long$time=rep(c(2, 3, 5, 8), rep(nobs, 4))


BtheB_lm=lm(bdi ~ bdi.pre + time + treatment + drug + 
              length, data=BtheB_long, na.action=na.omit)

summary(BtheB_lm)

BtheB_lmer1=lmer(bdi ~ bdi.pre + time + treatment + drug +
                   length + (1 | subject), data=BtheB_long,
                 REML=FALSE, na.action=na.omit)

cftest(BtheB_lmer1)

cat("Comparison of lm and lmer")
anova(BtheB_lmer1,BtheB_lm)


cat("We can plot the residuals from Linear model")
plot(BtheB_lm, which=2)
# Assumptions for Mixed-model

residuals <- function(object, obs) obs-predict(object) 
cat("We can also plot the residuals from linear mixed-model")
layout(matrix(1:2, ncol = 2))
qint <- ranef(BtheB_lmer1)$subject[["(Intercept)"]]
qres <- residuals(BtheB_lmer1,BtheB_long$bdi.pre)
qqnorm(qint, ylab = "Estimated random intercepts",
     xlim = c(-3, 3), ylim = c(-20, 20),
     main = "Random intercepts")
qqline(qint, col="red", lwd=3)
qqnorm(qres, xlim = c(-3, 3), ylim = c(-20, 20),
     ylab = "Estimated residuals",
     main = "Residuals")
qqline(qres, col="red", lwd=3)



cat("#1.d")
# d. Investigate and discuss whether there is any evidence of an interaction
# between treatment and time for the Beat the Blues data.

cat("Testing fitted linear model with lm")
BtheB_lm2=lm(bdi ~ bdi.pre + time + treatment + drug +
                   length + time*treatment + subject, 
                 data=BtheB_long, REML=FALSE, na.action=na.omit)
summary(BtheB_lm2)

cat("Testing fitted linear mixed-model with lmer subject as a random effect")
BtheB_lmer2=lmer(bdi ~ bdi.pre + time + treatment + drug +
                   length + time*treatment + (1 | subject), 
                 data=BtheB_long, REML=FALSE, na.action=na.omit)

cftest(BtheB_lmer2)

cat("#1.e")
# e. Construct a plot of the mean profiles of both treatment groups in the Beat
# the Blues data, showing also standard deviation bars at each time point.

data_sum=ddply(BtheB_long, c("time", "treatment"), summarise, 
            N=sum(!is.na(bdi)),
            mean=mean(bdi, na.rm=TRUE),
            sd=sd(bdi, na.rm=TRUE),
            se=sd/sqrt(N))

pd <- position_dodge(0.1)

ggplot(data_sum, aes(x=time, y=mean, colour=treatment)) +
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=0.1, position=pd) +
  geom_line(position=pd) +
  geom_point(position=pd) + 
  theme_classic() +
  labs(title = "Mean profiles of both treatment groups")



```
  
  Discussions:
  
  1a.
  
The plots demonstrate that BDI (median or average) declines at a faster rate they do not take the medication.
  
  
  1b. 
  
  The plots demonstrate that when patients live for less than 6 months, BDI (median or average) declines at a higher pace than when patients survive for more than months.
  
  1c.
  
  In linear model, Bdi.pre, time, treatmentBtheB, and drugYes are significant, but in the lmer model, only bdi.pre and time are significant. This means that the linear model understates the standard errors because it does not account for the correlation over time or the repeated measures.
  
  1d.
  
 In the first output, the linear model shows that there is interation between time and treatment with P < 0.05. However, mixed effect model (lmer) indicates that there is no time and treatment effect because p is 0.09.


  1e. 

  The mean profile shows that the results are very different between the two categories, but the standard deviation bars converge.
  
  
2. Consider the \textbf{phosphate} data from the package HSAUR3. This data shows the plasma inorganic phosphate levels for 33 subjects, 20 of whom are controls and 13 of whom have been classified as obese (Davis, 2002). Perform the following on this dataset
 
      a. Construct boxplots by group and discuss. 
   
      b. Produce separate plots of the profiles of the individuals in each group.
    
      c. Guided by how these plots fit, which linear mixed effects models do you think might be sensible? (Hint:                Discuss intercept and slope, intercept and interaction).
    
      d. Convert the data to long version and fit the model of your choice and discuss the results. 
      
```{r}
data("phosphate", package = "HSAUR3")
library(lme4)
library(Matrix)
library(multcomp)
attach(phosphate)
par(mfrow = c(1,2))
ylim <-range(phosphate[,-1], na.rm = TRUE)
cat("#2.a")
# a. Construct boxplots by group and discuss. 
boxplot(phosphate[group == "control", -1], 
        xlab = "time", ylab = "Phosphate level",
        main = "base R: Control group", ylim = ylim)
boxplot(phosphate[group == "obese",-1],
        xlab = "time", ylab = "Phosphate level",
        main = "Obese group", ylim = ylim)

# ggplot version:

phosphate %>%
  gather(time,phosphate_level, 2:9) %>%
  mutate(time = factor(gsub(time,pattern="[[:alpha:]]",replacement=''))) %>%
ggplot(aes(x=time,y=phosphate_level)) +
  geom_boxplot() +
  facet_grid(.~group) +
  labs(title='ggplot: Phosphate Data by Groups',
       x='Time (hours)',
       y='Phosphate level') +
  theme_classic()

cat("#2.b")
# b. Produce separate plots of the profiles of the individuals in each group
layout(matrix(1:2, nrow=1))

rg=range(phosphate[ ,-1])
plot(c(1, 8), c(rg), type="n", xlab="Time (hours)", 
     ylab="Phosphate Levels", main="base R: Control Group")
for(i in 1:sum(phosphate$group=="control")){
  lines(1:8, phosphate[i, -1])
}

plot(c(1, 8), c(rg), type="n", xlab="Time (hours)", 
     ylab="Phosphate Levels", main="Obese Group")
for(i in 1:sum(phosphate$group=="obese")){
  lines(1:8, phosphate[nrow(phosphate)-i, -1])
}


phosphate_t <- phosphate
phosphate_t$subject <- factor(rownames(phosphate_t))

phosphate_t %>%
  gather(time,phosphate_level, 2:9) %>%
  mutate(time = factor(gsub(time,pattern="[[:alpha:]]",replacement=''))) %>%
ggplot(aes(x=time,y=phosphate_level)) +
  geom_line(aes(group=subject,color=subject),size=1) +
  facet_grid(.~group) +
  labs(title='ggplot: Phosphate by group',
       x='Time (in hours)',
       y='Phosphate Level') +
  theme_classic()

cat("#2.d") 
#d. Convert the data to long version and fit the model of your choice and
#discuss the results.
data(phosphate)
ph=phosphate
ph$subject=as.factor(rownames(ph))
##LOng form
a=ph[,c(3:9)]

ph_long=reshape(ph,idvar="subject",v.names="p_level",
                varying=list(names(a)),direction="long")

timer=seq(1:7)
timer[1]=0.5
timer[2]=1
timer[3]=1.5
for(i in 4:7){
  timer[i]=i-2}

for(i in 1:7){
  for(j in 1:33){
    ph_long[(33*(i-1)+j),4]=timer[i]
  }
}

ph_non_inter=lmer(p_level~time+group+t0+(1|subject),data=ph_long,
                  REML=FALSE,na.actio=na.omit)

ph_inter=lmer(p_level~time+group+t0+time*group+(1|subject),
              data=ph_long,REML=FALSE,na.actio=na.omit)

ph_cept_slope=lmer(p_level~time+group+t0+
                     (time|subject),
              data=ph_long,REML=FALSE,na.actio=na.omit)

ph_cept_slope_inter=lmer(p_level~time+group+t0+time*group+
                           (time|subject),
                         data=ph_long,REML=FALSE,na.actio=na.omit)

print(summary(ph_cept_slope_inter))
print(cftest(ph_cept_slope_inter))
anova(ph_inter,ph_non_inter,ph_cept_slope,ph_cept_slope_inter)

pred=predict(ph_cept_slope_inter)
MSE=mean((pred-ph_long$p_level)^2)
message("Mean Squared Error: ",MSE)
message("Mean Phosphate Level: ",mean(ph_long$p_level))
```
   
   
   2.a
   
  Based on the plot, the control group has the median level of phosphate falling rapidly from a higher level of 4.3 to midway between t1 and t2 and starts to rise more rapidly over time. The obese group reached a relatively higher median phosphate rate close to 5 and dropped and grew steadily over time. The Obese group also has two outliers.


2.c

  I think linear mixed model would be suitable because the slopes and intercepts in each group vary widely. Additionally, there is an interaction between control and obese groups based on time. So fitting something of the form lmer(phosphatelevel t0+time∗group+(1= phos) would be suitable.


2.d


  Based on the model I used:
In the fixed effect we see that there is a positive relation between the time and the phosphate level. It makes sense sinse we saw that after decreasing for a while the phosphate level increases for longer than decreasing time period.  Based on the f-test with cftest  , time, group, baseline phosphate level and the interaction between time and group all are significant. 

Comparing this with three other models, (ph_non_inter=random intercept, ph_inter=random intercept with interction, ph_cept_slope=random intercept and random slope) using anova, the model I used has the smallest AIC (about 391) indicating a better fit of among four models. However, considering BIC value, we also see that random intercept with interaction is also a good model.
